---
title: "Working with LISST data in R"
author: "Alexandre Castagna"
output:
  html_document:
    toc: true
    theme: united
bibliography: measurement_units_in_R.bib
vignette: >
  %\VignetteIndexEntry{Working with LISST data in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Introduction

This vignette describes the use of the lisst package to read, process and 
visualize data recorded by the Laser In-Situ Scattering and Transmissometry 
(LISST) instruments, produced by Sequoia Scientific, Inc. The package was 
created to have a single frame work of functions and structures to work with 
different LISST models, simplifying analysis task when dealing with multiple
instruments or upgrade to a new model. The [oce](https://cran.r-project.org/package=oce) 
pakage provides function for visualition of LISST data, but is limited to the 
LISST-100(X) and to processed data (particle size distribution) as produced by 
the LISST-SOP and don't include specific processing or subsetting utilities.

## Quick overall demonstration

The following chapters will better describe the functionalities, but here is a 
quick overview from first time registering an instrument to the final plot of the
desired quantity.

```{r}
# Register the LISST instrument:
#path  <- system.file("extdata", package = "lisst")
#model <- 100
#lisst_reg(model, path)

# Read a LISST data file:
#sn    <- 1298
#pl    <- 0.05
#yr    <- 2018 
#zscat <- system.file("extdata", "bg_20180326.asc", package = "lisst")
#flp   <- system.file("extdata", "DN_27_rs.asc", package = "lisst") # LISST-SOP processed file
#flb   <- system.file("extdata", "DN_27.DAT", package = "lisst")    # LISST binary file

#lop   <- read_lisst(flp, sn, pl, zscat, yr)
#lob   <- read_lisst(flb, sn, pl, zscat, yr)

# Subset by sample, deth or time:
#lob_ss <- lob[1:3, ]         # First three samples
#lob_ss <- lob['0|5', ]       # First five meters
#lob_ss <- lob['2018-03', ]   # Samples in march 2018

# Conversion between data types:
#lop_pnc <- lget(lop, 'pnc')
#lob_cal <- lget(lob, 'cal')

# Concatenation:
#lop_d <- c(lop, lop_pnc)

# Statistics:
#lob_mn <- lgetstat(lob, mean)
#lob_md <- lgetstat(lob, media)
#breaks <- c(0, 5, 10, 15)
#lob_mn <- lgetstat(lob, mean, s = breaks) # Breaks by sample index
#lob_mn <- lgetstat(lob, mean, d = breaks) # Breaks by sample depths
#breaks <- c("2018-01", "2018-02", "2018-03", "2018-04")
#lob_mn <- lgetstat(lob, mean, t = breaks) # Breaks by sample time
 
# Plots:
# par(mfcol = c(2, 2))
# plot(lob[40:144, ])
# plot(lop[40:144, ], type = 'pnc')
# plot(lob[40:144, ], type = 'pf', xu = 'degree')
# plot(lop[40:144, ], type = 'vol')
```

## Reading data

The new models of the LISST instrument keep track of the instrument information
(e.g., model, instrument calibration constants) in the recorded data, but the 
older LISST-100(X) models don't. The lisst package reproduces the behaivour of 
the new models, by first registering the instrument specific information in an 
internal data base with the function `lisst_reg`, and by adding all relevant 
information in sperate attributes of the lisst object uppon read with the 
`read_lisst` function. In that way, only the instrument serial number has be 
informed for the `read_lisst` function to know which instrument it is working 
with. And since all information is stored in the object, additional processing 
deals only with the object. Here is an example on how to register a LISST, for the 
LISST-100X SN1298:

```{r}
#path  <- system.file("extdata", package = "lisst")
#model <- 100
#lisst_reg(model, path)
```

The registration need only to be done once, and is saved to the package data, 
such that it will be available for any R session in the same computer. The 
LISST-200X can be registered in the same way, but argument path must be the path
to a binary data file from which the informations will be extracted. Once the 
LISST is registered, reading an data file is simple and can be performed in 
diferent ways:

```{r}
#sn    <- 1298
#pl    <- 0.05
#yr    <- 2018 
#zscat <- system.file("extdata", "bg_20180326.asc", package = "lisst")

# For a processed file:
#flp   <- system.file("extdata", "DN_27_rs.asc", package = "lisst")
#lop   <- read_lisst(flp, sn, pl, zscat, yr)

# For a binary file:
#flb   <- system.file("extdata", "DN_27.DAT", package = "lisst")
#lob   <- read_lisst(flb, sn, pl, zscat, yr)
```

As the example shows, the function will determine the file type automaticly, 
based on its extension. In the example, the path length (pl, m) used and the year 
of first measurement (yr, required for LISST-100(x)) are provided, besides the
background data file path (zscat). The first two can be omitted, and, with a 
warning, the function will use the standard path length for the instrument and 
apply a 'best guess' for yr (read the simple logic for this in the documentation 
of `lgdate`). The background file can also be omitted when reading a processed 
file or when reading a binary file when the requested output is raw digital counts
(out = 'raw'). In the last case, however, further processing of the data (e.g., 
to VSF) will not be possible. Another variation that is worth noting is how to read 
a file for an instrument that is not registered - this is limited to the same 
cases when zscat can be omitted, and instead of providing sn, a model is sufficient.

```{r}
#lom <- read_lisst(flp, pl = pl, yr = yr, model = "100CX")
```
As a final note, when reading a processed file or when reading a binary file as 
calibrated units (out = 'cal', the default for binary files) the lisst object will
have an additional column, with date/time in POSIXct format.

## The lisst object

The lisst object is a `data.frame` in wich the variables are of class `units`, and 
with attributes that hold information on the data and instrument. Those attributes
are not expected to be manipulated directly by the user. It has essentially the
same structure as the output of the LISST-SOP, except for the Time column. There 
is, however, a difference in the data read by `read_lisst` from the one produced 
by the LISST-SOP: uppon reading, the optical transmission and the beam 
attenuation are corrected to add effect of pure water. A correction is also
applied in the rings digital counts for the binary data.

```{r}
#lop[15, , drop = FALSE]
#lob[15, , drop = FALSE]
```

## Processing

As of this version, is not yet possible to directly invert the PSD from the 
scattering data in the binary files. But is possible to convert between 
representations of the PSD (volume concentration x particle number concentration),
different levels of processing of the binary data (digital counts x corrected 
digital counts x calibrated values) and retrieving the particle Volume Scattering 
Function form the calibrated values.

```{r}
# Converting from vol conc to n conc:
#lop[15, 1:3, drop = FALSE]
#lgetpnc(lop)[15, 1:3, drop = FALSE]

# Converting from cal to VSF:
#lob[15, 1:3, drop = FALSE]
#lgetvsf(lob)[15, 1:3, drop = FALSE]
```

Is also possible to extract PSD model parameters. At the curent version, only the
Junge model is available:

```{r}
# Extract the Junge slope for a PSD:
#lgetfit(lop)[15:17]
```

## Visualizing

The visualization use the common R plot function, with different plot types 
dependent on lisst object type. The units used in the axis can also be modified, 
but note that unit conversion is performed automaticly by the units package, such 
that if the original data units cannot be converted into the requested units, 
it will generate an error. Here is an example on the visualization of the particle 
VSF, the approximate particle phase function (normalization by c, not b), the number
concentration and the volume concentration.

```{r}
#lov <- lgetvsf(lob)
#par(mfcol = c(2, 2))
#plot(lov[40:144, ])
#plot(lop[40:144, ], type = 'pnc')
#plot(lov[40:144, ], type = 'pf', xu = 'degree')
#plot(lop[40:144, ], type = 'vol')
```
